---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


REPRODUCTIBLE RESEARCH FIRST PEER ASSESSMENT
========================================================

The data consists of two months of data collected through  devices such as a Fitbit, Nike Fuelband, or Jawbone Up, from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Load and Procesing the Data

### Load the table

```{r}
setwd(dir="~/Desktop/Reproductible research")
activity<- read.csv(file="activity.csv", header=TRUE, sep=",", fill= TRUE)
```

let s see what kind of variables are we dealingh with

```{r}
str(activity)
summary(activity)
```

### Procesing the data

1.Steps is an intergres. As it counts steps we can keep this variable as it is for further analysis. 
2.Date is coded as a factor. Probably, as in the assigment it is asked to create an histogram by date, this should be change as a date. 

```{r}
activity$date<- as.Date(activity$date,"%Y-%m-%d")
```

3. Interval should be a factor, as it tells which interval has been used to record the data

```{r}
activity$interval<- as.factor(activity$interval)
```

let's see if changes has been performed correctly

```{r}
str(activity)
```

## What is mean total number of steps taken per day?

As the assigment says 
"For this part of the assignment, you can ignore the missing values in the dataset".
I will keep the "NA" values.

## Calculate the total number of steps taken per day

The idea would be to aply the summary or average function to the table subseting by "date"
First I need the function "ddply" in the "plyr" package. So I load the package. 

```{r}
library("plyr", lib.loc="/Library/Frameworks/R.framework/Versions/3.1/Resources/library")
stepaday<-ddply (activity, .(date), summarize, tot= sum(steps, na.rm=TRUE))
```

now, we use the new calculated value to draw the histogram

```{r}
hist(stepaday$tot,main = "Total steps by day", xlab = "day", col="blue")
```

to report the mean and the median of steps for each day, I can use again "ddply"

```{r}
avg<-ddply (activity, .(date), summarize, tot= sum(steps, na.rm=TRUE), median= median(steps, na.rm=FALSE),mean= mean(steps, na.rm=FALSE))
```

##What is the average daily activity pattern?

##Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

we already know that "interval" is a factor. 
we could apply the same formula to calculate the average number of steps each interval.

```{r}
interval<-ddply(activity,.(interval), summarize, avg= mean (steps, na.rm= TRUE))
plot(interval$avg, type="l", col="blue", main= "avg steps", xlab="interval", ylab="steps")
summary(interval)
```

to calculate the interval with the maximum number of steps, I apply the "which.max" function to the table just created

```{r}
which.max(interval$avg)
```

once I have the position "104", I ask what is the value "104" in the table

```{r}
interval$interval[104]
```

## Imputing missing values
## Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

First, I figure out which values are "NA" in the "steps" column and create a new vector

```{r}
na<- which(is.na(activity$steps))
```

Second, I ask for the lenght of the vector na

```{r}
length(na)
```

## Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.


First I duplicate the activity table into steps

```{r}
test<-activity
```

I perform a loop afterwards to change each NA value by the average steps value for the interval. 

```{r}
for (n in 1:nrow(test))
{ 
 if (is.na(test[n,1])) {
       test$steps[n] <- interval$avg[which(interval$interval== test[n,3])]
          } 
}
```

to check the result

```{r}
head(test)
```

## Create a new dataset that is equal to the original dataset but with the missing data filled in

test is the dataset already created. 

## Make a histogram of the total number of steps taken each day 

Again, first I apply "ddply" function to calculate the new total number of steps each day

```{r}
stepaday1<-ddply (test, .(date), summarize, tot= sum(steps, na.rm=TRUE))
head(stepaday1)
```

now, the histogram

```{r}
hist(stepaday1$tot,main = "Total steps by day", xlab = "day", col="green")
```

## Calculate and report the mean and median total number of steps taken per day

```{r}
avg1<-ddply (test, .(date), summarize, tot= sum(steps, na.rm=TRUE), median= median(steps, na.rm=FALSE),mean= mean(steps, na.rm=FALSE))
```

to compare the values, I can merge both tables (mean and mean1) by rows creating an new table

```{r}
avg2<- merge.data.frame(avg, avg1, by="date")
head(avg2)
```

it is clear the values are different. The "NA" values create new "NA" values in the logical operations. 


## Are there differences in activity patterns between weekdays and weekends?

## Create a new factor variable in the dataset with two levels ??? ???weekday??? and ???weekend??? indicating whether a given date is a weekday or weekend day.

I create the new variable in the table "test"

```{r}
test2<- as.factor(weekdays(test$date))
test3<- test
test3$day<- as.character(test2)
```

Now create a new variable for weekday or weekend. For changing the approach, 
First I create a vector with the weekend vaules

```{r}
Weekend <- which(test3$day == "Saturday"|test3$day =="Sunday")
```

Second I assign a new variable with the value "Weekday"

```{r}
test3$week<- "Weekday"
```

Third, I change the value "Weekday" for "Weekend" for the weekend days

```{r}
test3$week[Weekend]<- "Weekend"
```

I make sure the variable is a factor and check the levels

```{r}
test3$week<-as.factor(test3$week)
levels(test3$week)
```

## Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
##See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
interval3 <- aggregate(steps ~ interval + week, data = test3, mean)
library(lattice)
xyplot(steps ~ interval | week, interval3, type = "l", layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")
```





